NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-04-26 Manuscript/manuscriptPS.tex
Update manuscript and figures. Add Berger data for PC change as a function of PS.

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Order parameters of PC (left) and PS (right) headgroup as a function of
    added NaCl, KCl and LiCl measured from POPC:POPS (5:1) mixture \cite{roux90}.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should get the CHARMM results as well?}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Headgroup order parameters from PC:PS mixtures from 
    different simulation models and experiments.
    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
    signs are determined as discussed in \cite{botan15,ollila16}).
    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
    Counterions in expriments are sodium, while potassium is used in simulations.
    Using sodium in simulations do not have a significant effect.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should show the changes also in this plot?}
\end{figure*}

@@

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) bilayers
as a function of Ca$2+$ concentration from different
simulations and experiments \cite{roux90} are shown in Fig. \ref{changesWITHCaClPS}.
The results suggest that Ca2+ ions clearly overbind in simulations with
MacRog model, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments \todo{Should be analyze/discuss this further?}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, which are incorporated
to CHARMM-GUI. These parameters give also too weak binding of calcium to pure POPC bilayer,
as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
The density profiles of Ca2+ from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
\todo{More simulation data for systems with PS lipids and CaCl$_2$ would be probably useful}

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    POPC (left) and POPS (right) headgroup order parameters in lipid bilayer with POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations.
  }
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, numbers should be used.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Not all the data from MacRog is included.} 
\end{figure}

Also the order parameters of PS headgroup as a function of CaCl$_2$ concentration
%from mixtures with PC  
are shown in Fig. \ref{changesWITHCaClPS}. The order parameters exhibit
a strong dependence small CaCl$_2$ concentrations with a rapid saturation. 
The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
both simulation models. The response of $\beta$-carbon order parameter could be
interpreted qualitatively correct in simulations, but the general agreement with
experiments is not sufficient to use simulations to interpret the response of
PS headgroup to the added CaCl$_2$.

@@

\subsection{Berger}
\todo{To be wiritten by Piggot and Ollila}
Simulations with sodium were taken directly from Ref. \citenum{??} and
simulations with calcium directly from \citenum{melcrova16}.

>>>>

\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
\affiliation{nmrlipids.blogspot.fi}

@@

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{Maybe we should get the CHARMM results as well?}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
\end{figure*}

@@

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) mixtures
are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
simulations and experiments \cite{roux90}. 
The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments
\todo{Should be analyze/discuss this further?
  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
It should be noted that Berger simulation do not have counterions.}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
The binding of calcium to pure POPC bilayer is also too weak with these parameters
as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
\todo{The discussion is to be finished when we have all the data in the plot.}

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero. 
  }
  \todo{Information about the cuonterions in different simulations should be added} \\
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, numbers should be used.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Not all the data from MacRog is included.} 
\end{figure}

Also the order parameters of PS headgroup from POPC:POPS (5:1) mixture
are shown in Fig. \ref{changesWITHCaClPS} as a function of CaCl$_2$ concentration.
In experiments the order parameters exhibit a strong dependence of CaCl$_2$ with
small concentrations with a rapid saturation around 50 mM. 
The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
all tested simulation models. Furthermore, the changes of the headgroup order parameters
do not qualitatively agree with experiments. This is in contrast to previous results
for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
observed despite of significant discrepancies in the headgroup structure without additional ions.

@@

\subsection{Berger}
\todo{To be wiritten by Piggot and Ollila}
Simulations with sodium were taken directly from Ref. \citenum{??} and
simulations with calcium directly from \citenum{melcrova16}.
Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.
